---
title: "class 15 whooping cough"
author: "Hannah Lewack"
format: pdf
editor: visual
---

```{r}
library(datapasta)
cdc<-data.frame(Year = c(1922L,1923L,1924L,1925L,1926L,1927L,1928L,
                    1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                    1936L,1937L,1938L,1939L,1940L,1941L,1942L,
                    1943L,1944L,1945L,1946L,1947L,1948L,1949L,
                    1950L,1951L,1952L,1953L,1954L,1955L,1956L,
                    1957L,1958L,1959L,1960L,1961L,1962L,1963L,
                    1964L,1965L,1966L,1967L,1968L,1969L,1970L,
                    1971L,1972L,1973L,1974L,1975L,1976L,1977L,
                    1978L,1979L,1980L,1981L,1982L,1983L,1984L,
                    1985L,1986L,1987L,1988L,1989L,1990L,1991L,
                    1992L,1993L,1994L,1995L,1996L,1997L,1998L,
                    1999L,2000L,2001L,2002L,2003L,2004L,2005L,
                    2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                    2013L,2014L,2015L,2016L,2017L,2018L,2019L,
                    2020L,2021L,2022L),
No..Reported.Pertussis.Cases = c(107473,164191,165418,152003,
                                 202210,181411,161799,197371,
                                 166914,172559,215343,179135,
                                 265269,180518,147237,214652,
                                 227319,103188,183866,222202,
                                 191383,191890,109873,133792,
                                 109860,156517,74715,69479,
                                 120718,68687,45030,37129,60886,
                                 62786,31732,28295,32148,40005,
                                 14809,11468,17749,17135,13005,
                                 6799,7717,9718,4810,3285,4249,
                                 3036,3287,1759,2402,1738,1010,
                                 2177,2063,1623,1730,1248,1895,
                                 2463,2276,3589,4195,2823,3450,
                                 4157,4570,2719,4083,6586,4617,
                                 5137,7796,6564,7405,7298,7867,
                                 7580,9771,11647,25827,25616,
                                 15632,10454,13278,16858,2755,
                                 18719,48277,28639,32971,2076,
                                 17972,18975,15609,18617,6124,
                                 2116,3044))
```

general plot of whopping cough cases by year

```{r}
library(ggplot2)
ggplot(cdc) +
  aes(x=Year, y=No..Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  labs(x="Year", y= "Number of Cases", title="Pertussis Cases by Year (1922-2019)")
```

plot of whopping cough cases with the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine

```{r}
ggplot(cdc) +
  aes(x=Year, y=No..Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  labs(x="Year", y= "Number of Cases", title="Pertussis Cases by Year (1922-2019)")+
geom_vline(xintercept=1946, linetype = "dotted", color = "blue") +geom_vline(xintercept = 1996, linetype = "dotted", color = "red")+geom_vline(xintercept = 2020, linetype = "dotted", color = "green")+ annotate("text", x = 1946, y = max(cdc$No..Reported.Pertussis.Cases), label = "WP", color = "blue", vjust = -0.5) +
  annotate("text", x = 1996, y = max(cdc$No..Reported.Pertussis.Cases), label = "AP", color = "red", vjust = -0.5)+annotate("text", x = 2020, y = max(cdc$No..Reported.Pertussis.Cases), label = "COVID", color = "green", vjust = -0.5)
```

-   **Q3.** Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

    after the AP vaccine numbers started to go up, is this because of the anti vax movement? in 1998 someone first came up with the claim that the vaccine causes autism which is not far after 1996

    ```{r}
    library(jsonlite)
    library(dplyr)
    subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
    head(subject, 3)
    ```

<!-- -->

-   **Q4.** How many aP and wP infancy vaccinated subjects are in the dataset?

    ```{r}
    subject%>%filter(infancy_vac=="wP")%>%count()
    subject%>%filter(infancy_vac=="aP")%>%count()
    ```

**Q5.** How many Male and Female subjects/patients are in the dataset?

```{r}
subject%>%filter(biological_sex=="Female")%>%count()
subject%>%filter(biological_sex=="Male")%>%count()
```

-   **Q6.** What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
table(subject$dataset)
#this does not reflect breakdown of us pop, These data are still useful, however.
```

-   **Q8.** Determine the age of all individuals at time of boost?

    ```{r}
    library(lubridate)
    int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
    age_at_boost <- time_length(int, "year")
    head(age_at_boost)
    ```

    switching subjects

    ```{r}
    specimen <- read_json("http://cmi-pb.org/api/v5/specimen",
                          simplifyVector = TRUE)
    ab_titer <- read_json("http://cmi-pb.org/api/v5/plasma_ab_titer",
                          simplifyVector = TRUE)
    head(specimen)
    head(ab_titer)
    library(dplyr)
    meta <- inner_join(subject, specimen)
    head(meta)
    ```

    ```{r}
    abdata <- inner_join(ab_titer, meta)
    head(abdata)
    nrow(abdata)
    table(abdata$isotype)
    table(abdata$antigen)
    ```

    looking at igg, Make a boxplot of IgG antigen levels. This will be a plot of MFI vs antigen.

    ```{r}
    igg <- filter(abdata, isotype=="IgG")
    head(igg)
    ggplot(igg, aes(MFI_normalised, antigen)) +
      geom_boxplot()
    ggplot(igg, aes(MFI_normalised, antigen, color = infancy_vac)) +
      geom_boxplot()
    ```

Ideally, we would like to see how these Ab levels change over time relative to the booster shot.

```{r}
table(abdata$visit)
ggplot(igg)+
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
ggplot(igg)+
  aes(MFI_normalised, antigen, color = infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit)
iggpt <- filter(igg, antigen=="PT")
head(iggpt)
iggpt21 <- filter(iggpt, dataset=="2021_dataset")
head(iggpt21)
```

2021 dataset igg

```{r}
ggplot(iggpt21) +
  aes(x= planned_day_relative_to_boost,
      y= MFI_normalised,
      color = infancy_vac,
      group = subject_id)+
  geom_point()+
  geom_line()+
  geom_vline(xintercept=0, linetype="dashed")+
  geom_vline(xintercept=14, linetype="dashed")+
  labs(title="2021 Dataset IgG PT", subtitle="Dashed lines indicate day 0 (pre-boost) and day 14 (peak values")
```
